---
title: "DataExploration"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r}
library(Seurat)
library(SingleCellExperiment)
library(dplyr)
library(Matrix)
library(ggplot2)
```

```{r load the data}
load('/Users/ameyateli/LAs BeST/Single Cell/sce.Rdata')
```

use R to plot the distribution of a single gene across all the cells
pick diff genes, play around with plots etc. assay, functions in the
document try a few transformations to see something

```{r}
# exploring the data, 
class(sce)

Assays(sce)

dim(sce) # dimensions of count matrix

sce[1:10, 1:3] # examining matrix

sce[7, 1:3]
```

```{r}
counts <- counts(sce) # accessing the counts matrix


# looking at a specific cell/cells and its totals 
counts_per_cell <- Matrix::colSums(counts)

cat("counts per cell: ", counts_per_cell[1:5], "\n") ## counts for first 5 cells

# cell number 7
cat("counts per cell: ", counts_per_cell[7], "\n") ## counts for cell 7

```

```{r}
# looking at a specific gene and its totals 
counts_per_gene <- Matrix::rowSums(counts)

# gene number 7
cat("counts per gene: ", counts_per_gene[7], "\n") ## counts for cell 7


# barplot of row7: gene7
gene7 <- counts[7, ]

barplot(gene7, main = "Times Gene 7 Was Expressed Per Cell")

hist(gene7, main = "Frequency of Cells expressing gene 7", breaks=11, right = FALSE)
hist(log10(gene7), main = "Frequency of Cells expressing gene 7")

# this gene is called UBE2J2
jpeg(filename = "gene7_ube2j2_plot.jpg", width = 800, height = 600, quality = 90)
hist(gene7, main = "Frequency of Cells expressing UBE2J2", breaks=11, right = FALSE, col = "mediumblue", xlab = "Gene UBE2J2")
dev.off()


```

```{r}

# barplot of row 47: gene 47
gene47 <- counts[47, ]

barplot(gene47, main = "Times Gene 47 Was Expressed Per Cell")

hist(gene47, main = "Frequency of Cells expressing gene 47", breaks=11, right = FALSE)
hist(log10(gene47), main = "Frequency of Cells expressing gene 47")

boxplot(gene47, main = "Frequency of Cells expressing gene 47")

```

```{r}
# both gene 7 and gene 47
boxplot(gene7, gene47, main = "Frequency of Cells expressing gene 7 and gene 47")
```

```{r}
# testing different genes 
hist(counts[6000,])


# histogram of MT-ATP6
curr_row <- counts[c("MT-ATP6"),]
hist(curr_row, right = FALSE, breaks = max(curr_row)+1, main = "Frequency of Cells expressing MT-ATP6")
```

```{r}
# exploring metadata
cell_info <- colData(sce)
head(cell_info)
```

```{r}
# 
hist(log(gene7 + 1), main = "Frequency of Cells expressing gene 7", breaks=11, right = FALSE)
hist(log(gene7), main = "Frequency of Cells expressing gene 7", breaks=11, right = FALSE)


# Load the package
library(scater)
```

```{r}
plotColData(sce, y = "nGene")
plotColData(sce, y = "stage")
plotColData(sce, y = "type")
plotColData(sce, y = "nUMI")
plotColData(sce, y = "doublet_score")
plotColData(sce, x = "sample", y = "type")

colData(sce, y = "sample")

plotColData(sce, x = "sample", y = "stage")


```

```{r}
colnames(colData(sce))
```

```{r}
plotColData(sce, x = "stage", y = "nGene", 
            color_by = "stage") + 
  scale_color_manual(values = c(
    "early" = "darkolivegreen", 
    "locAdv" = "gold", 
    "metast" = "tomato"
  ))
```

```{r}
# normalization 

sce <- logNormCounts(sce)
assayNames(sce)

log_counts <- logcounts(sce) # accessing the counts matrix

# after normalization 
log_gene47 <- log_counts[47,]
hist(log_gene47, main = "Frequency of Cells expressing gene 47", breaks=11, right = FALSE)

# before normalization to compare 
gene47 <- counts[47, ]
hist(gene47, main = "Frequency of Cells expressing gene 47", breaks=11, right = FALSE)

# gene 47 is called SPEN 
jpeg(filename = "gene47_spen_plot.jpg", width = 800, height = 600, quality = 90)
hist(gene47, main = "Frequency of Cells expressing SPEN", breaks=11, right = FALSE, col = "maroon", xlab = "Gene SPEN")
dev.off()

# plotting them together 
par(mfrow = c(1, 2)) 
hist(log_gene47, main = "Normalized Frequency of Cells expressing gene 47", breaks=11, right = FALSE, col = "mediumpurple2")
hist(gene47, main = "Frequency of Cells expressing gene 47", breaks=11, right = FALSE, col = "hotpink4")
```



```{r}
# trying this on mtatp 
log_curr_row <- log_counts[c("MT-ATP6"),]
hist(log_curr_row, right = FALSE, breaks = max(log_curr_row)+1, main = "Frequency of Cells expressing MT-ATP6")
curr_row <- counts[c("MT-ATP6"),]
hist(curr_row, right = FALSE, breaks = max(curr_row)+1, main = "Frequency of Cells expressing MT-ATP6")

#normalization works!!! 
jpeg(filename = "normalization_plots.jpg", width = 800, height = 600, quality = 90)
par(mfrow = c(1, 2), oma = c(0, 0, 3, 0)) 
hist(curr_row, right = FALSE, breaks = max(curr_row)+1, main = "Before Normalization", col = "magenta3", border = "magenta3", xlab = "Gene Expression Counts")
hist(log_curr_row, right = FALSE, breaks = max(log_curr_row)+1, main = "Normalized Frequency of MTATP6", col = "olivedrab2", xlab = "Log of Gene Expression Counts")
mtext("Frequency of Gene Expression Counts for MTATP6", side = 3, outer = TRUE, cex = 1.5, font = 2, line = 1)
dev.off()

jpeg(filename = "before_norm.jpg", width = 800, height = 600, quality = 90)
hist(curr_row, right = FALSE, breaks = max(curr_row)+1, main = "Before Normalization", col = "magenta3", border = "magenta3", xlab = "Gene Expression Counts")
dev.off()

hist(curr_row, right = FALSE, breaks = max(curr_row)+1, main = "Before Normalization", col = "magenta3", border = "magenta3", xlab = "Gene Expression Counts")
hist(log_curr_row, right = FALSE, breaks = max(log_curr_row)+1, main = "Normalized Frequency of MTATP6", col = "olivedrab2", xlab = "Log of Gene Expression Counts")

```

## Using a Seurat Object 
-following the tutorial on the seurat website, a guided clustering tutorial 

```{r}
# trying the second option - converting SingleCellExperiment object to a Seurat object 
sce_meta <- as.data.frame(colData(sce))

# creating seurat object from single cell experiment counts matrix
seurat_obj <- CreateSeuratObject(
  counts = counts, 
  meta.data = sce_meta, 
  project = "sce")

# normalizing the seurat object
norm_so <- NormalizeData(seurat_obj)

norm_so
```

```{r}
# exploring the seurat object 

norm_so[1, 1:3]
norm_so

# looking at first 6 rownames
head(x = rownames(x = norm_so))

# looking at first 6 colnames
head(x = colnames(x = norm_so))

norm_so[['RNA']]

```

```{r}
# creating plots from Seurat normalized object 

row47_seu <- LayerData(norm_so, assay = "RNA", layer = "counts")[47,]

par(mfrow = c(1,2))

hist(row47_seu, breaks=11, right = FALSE)

# comparing our seurat normalized to the SCE normalization 
hist(gene47, main = "Frequency of Cells expressing gene 47, SCE object", breaks=11, right = FALSE)

```

Visualization of normal vs tumor cell counts for data presentation
slideshow

```{r, eval= FALSE}
# plotColData(sce_meta, y = type)
# 
# cell_counts <- table(sce_meta$type)
# 
# barplot(cell_counts, xlab = "Normal vs. Tumor", ylab = "Number of Cells", col = c("skyblue", "tomato"))
# 
# jpeg(filename = "cell_count_normalvstumor.jpg", width = 800, height = 600, quality = 90)
# barplot(cell_counts, xlab = "Normal vs. Tumor", ylab = "Number of Cells", col = c("skyblue", "tomato"), main = "Cell Counts of Normal vs Tumor Samples")
# dev.off
```

6/24

normalization steps: divide by totals, colSums multiplies by 10,000 \<--
arbitrary large constant applies log normalization \<-- log1p(), does
log but adds one to ensure 0 still is 0, not -(infinity) \^natural log

follow seurat tutorial, get to pca and dimensionality reduction, then
elbow plot (pick PCA \#) if time UMAP \<-- clustering, just a function

Attempt at PCA using the Seurat object

norm_so is the seurat object that is normalized

### Identification of highly variable features (feature selection)

```{r}
norm_so <- FindVariableFeatures(norm_so, selection.method = "vst", nfeatures = 2000) 
# using the default number of features which is 2000

top10 <- head(VariableFeatures(norm_so), 10)
top10
```

```{r}
plot1 <- VariableFeaturePlot(norm_so)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot1 + plot2
```

Scaling the Data

```{r}
all.genes <- rownames(norm_so)
norm_so <- ScaleData(norm_so, features = all.genes)
```

Performing linear dimensional reduction

```{r}
norm_so <- RunPCA(norm_so, features = VariableFeatures(object = norm_so), seed.use = 10)
```

```{r}
print(norm_so[["pca"]], dims = 1:5, nfeatures = 5)
```

```{r}
VizDimLoadings(norm_so, dims = 1:2, reduction = "pca")
```

```{r}
DimPlot(norm_so, reduction = "pca") + NoLegend()
```

```{r}
DimHeatmap(norm_so, dims = 1, cells = 500, balanced = TRUE)
```

```{r}
ElbowPlot(norm_so, ndims = 50)
```

```{r}
set.seed(10)
norm_so <- FindNeighbors(norm_so, dims = 1:25)
norm_so <- FindClusters(norm_so, resolution = 0.5, random.seed = 10)
```

```{r}
norm_so <- RunUMAP(norm_so, dims = 1:25)
DimPlot(norm_so, reduction = "umap")
```

clusters number, make git hub doc, biology, bubble plot 



