---
title: "Single_Cell_Main_Code"
output: html_document
---

Some libraries we will need: 
```{r}
library(Seurat)
library(SingleCellExperiment)
library(dplyr)
library(Matrix)
library(ggplot2)
```

First we will load the data in: 
```{r}
# insert file path here 
load('/Users/ameyateli/LAs BeST/Single Cell/sce.Rdata')
```

We want to normalize the counts across the single cell experiment, to do this we can convert our single cell experiment object to a Seurat object to utilize the functionality built into Seurat: 
This website has useful information for these preliminary steps <https://satijalab.org/seurat/articles/pbmc3k_tutorial> 

## Normalization 
-following the tutorial on the seurat website, a guided clustering tutorial 

```{r}
counts <- counts(sce) # accessing the counts matrix

# converting SingleCellExperiment object to a Seurat object 

# first we need to extract the metadata from the single cell object
sce_meta <- as.data.frame(colData(sce))

# creating Seurat object from single cell experiment counts matrix
seurat_obj <- CreateSeuratObject(
  counts = counts, 
  meta.data = sce_meta, 
  project = "sce")

# normalizing the seurat object using NormalizeData 
norm_so <- NormalizeData(seurat_obj)

norm_so
```

The steps of normalization are as follows: 
- first each cell is divided by totals (colSums)
- next they are multiplied by an arbitrarily large constant, in this case it is 10,000
- now we apply log normalization, log1p is the function 
- log1p does log after adding 1, this is because of the amount of 0's in our data set 
- log reigns in outliers, makes data more normal, distribution after 0 

## Seurat Guided Clustering Tutorial 

Now we begin following the Seurat guided clustering tutorial. 

Step 1: Identification of highly variable features (feature selection)
We will select the first 2000 features, the 2000 most variable features (i.e. genes)

```{r}
norm_so <- FindVariableFeatures(norm_so, selection.method = "vst", nfeatures = 2000) 
# using the default number of features which is 2000

# look at the top 10 most variable features 
top10 <- head(VariableFeatures(norm_so), 10)
top10
```

We can also look at plots of the variable features, these plots include both the features of interest that we have selected as well as the other 4260 that aren't included. 

```{r}
plot1 <- VariableFeaturePlot(norm_so)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot1 + plot2
```

Step 2: Scaling the data 
This places the 'middle' of the data at the origin so we can do PCA. 
remove this 
```{r}
all.genes <- rownames(norm_so)
norm_so <- ScaleData(norm_so, features = all.genes)
```

Step 3: Performing linear dimensional reduction 
This code block below performs Principal Component Analysis (PCA)
```{r}
norm_so <- RunPCA(norm_so, features = VariableFeatures(object = norm_so), seed.use = 10)
```

View the 5 genes that have the highest coefficients with relation to the principal component. 
```{r}
print(norm_so[["pca"]], dims = 1:5, nfeatures = 5)
```

Viewing this graphically. 
```{r}
VizDimLoadings(norm_so, dims = 1:2, reduction = "pca")
```

Look at the PC1 and PC2 on a plot. We can see that the cells are more spread out on the PC1 axis which shows how this is the plane of the most variability. 
```{r}
DimPlot(norm_so, reduction = "pca") + NoLegend()
```

Looking at our elbow plot we can select the number of principal components which make sense for our analysis. We have selected 25 because that is where it appears to begin to decrease linearly, where most of the variability has been explained. 
```{r}
ElbowPlot(norm_so, ndims = 50)
```

Here we are finding the shared nearest neighbors of each of the cells only using the 25 PC's, we are currently in 25 dimensions. First we find the neighbors, then we find the clusters. We have set the seed for reproduce-ability.

Shared nearest neighbors works by creating an edge between two nodes if they share n amount of nearest neighbors. 
```{r}
set.seed(10)
norm_so <- FindNeighbors(norm_so, dims = 1:25)
norm_so <- FindClusters(norm_so, resolution = 0.5, random.seed = 10)
```

The following graph is running UMAP for clustering. This is not useful for downstream analysis due to the randomness of the UMAP algorithm. However, it is a helpful visualiztion for initial data exploration. 
```{r}
norm_so <- RunUMAP(norm_so, dims = 1:25)
DimPlot(norm_so, reduction = "umap")
```

## Data visualization 
<https://satijalab.org/seurat/articles/visualization_vignette>


```{r}
library(patchwork)
```

pbmc3k.final <- LoadData("pbmc3k", type = "pbmc3k.final")
features <- c("LYZ", "CCL5", "IL32", "PTPRCAP", "FCGR3A", "PF4")
pbmc3k.final

Features of interest from this website <https://www.cellsignal.com/pathways/immune-cell-markers-human?srsltid=AfmBOor-62g2ZCxcHumdrjyNY5RON2LIRvJ7sVe0PbQzU26i7AgghEVC> 
 HLA-DR, CD123, CD11c, HLA-DR, CD68, HLA-DR, CD11c, CD14, HLA-DR, CD206, CD86, CD3, CD56, CD3, CD56, CD3, CD19, BCMA, CD138

grep checks the data frame feature names (our genes) to see if they start with the value I have provided, this is helpful because the gene names from the website are not exactly what is in our data set 
```{r}
grep("CD200", rownames(counts), value = TRUE)
```


```{r}
features <- c("HLA-DRA", "HLA-DRB5", "HLA-DRB1", "CD68", "CD86", "CD3E", "CD3G", "CD3D", "CD1A", "CD5", "CD7", "CD200", "CD4", "CD8A", "CD8B", "CD1D", "CD160", "IL21R", "CD34", "CD38", "CD48", "CD14", "CXCR3", "CD1C", "CD163", "CD200R1", "CD81", "VEGFB", "VEGFA", "ZAP70", "CD207", "CD9", "CD19", "CD27", "CD33", "CD40", "CD69", "CD84", "FCER1A", "FLT3", "RORC", "KLRG1")

features <- sort(features)
```

```{r}
# Dot plots - the size of the dot corresponds to the percentage of cells expressing the
# feature in each cluster. The color represents the average expression level
DotPlot(norm_so, features = features) + RotatedAxis()
```
Percent expression is how many cells in that group have expressed that gene 
Average expression is the average expression of the gene among the cells in that cluster 

```{r}
norm_so$seurat_clusters

# 
norm_so$celltype <- as.character(norm_so$seurat_clusters)
norm_so$celltype

# example of how to label our cell clusters with types of cells
# norm_so$celltype[so$seurat_clusters == 1] <- 'CD8 Tcell'


```


### Fixing clusters to match group members 
Because my clusters were different than group members: I will read the clusters in and add them manually 

```{r}
clus <- readRDS('clusters.rds')
norm_so$seurat_clusters <- clus

#set the idents to be my new clustering 
Idents(norm_so) <- norm_so$seurat_clusters
```


Some features of interest that we wanted to vizualise on a bubble chart. 
```{r}
features <- c("HLA-DRA", "HLA-DRB5", "HLA-DRB1", "CD68", "CD86", "CD3E", "CD3G", "CD3D", "CD1A", "CD5", "CD7", "CD200", "CD4", "CD8A", "CD8B", "CD1D", "CD160", "IL21R", "CD34", "CD38", "CD48", "CD14", "CXCR3", "CD1C", "CD163", "CD200R1", "CD81", "VEGFB", "VEGFA", "ZAP70", "CD207", "CD9", "CD19", "CD27", "CD33", "CD40", "CD69", "CD84", "FCER1A", "FLT3", "RORC", "KLRG1", "ITGAM")

features <- sort(features)
```

```{r}
# Dot plots - the size of the dot corresponds to the percentage of cells expressing the
# feature in each cluster. The color represents the average expression level
DotPlot(norm_so, features = features) + RotatedAxis()
```
Percent expression is how many cells in that group have expressed that gene 
Average expression is the average expression of the gene among the cells in that cluster 

FindMarkers <-- function that tells you which genes differentiate 

### Making bubble plot that includes more features 

```{r}
m_features <-c("HLA-DRA", "HLA-DRB5", "HLA-DRB1","ITGAM", "ITGAX", "IL3RA","CD68", "CD14","CD86","FCGR3A","CLEC9A","CD1C","GPR56","SIRPA","KIT","FCER1A","CD163","BST2","SIGLEC1","CD83","THBD", "BCAM")

m_features <- sort(m_features)
```

```{r}
# Dot plots - the size of the dot corresponds to the percentage of cells expressing the
# feature in each cluster. The color represents the average expression level
DotPlot(norm_so, features = m_features) + RotatedAxis()
```

```{r}
l_features <-c("CD3D", "CD3E", "CD3G","CD8A","CD8B","CD4","CD19", "TNFRSF17","SDC1","SPI1","TBX21","IFNG","GATA3","FOXP3","TIFA","AHR", "CD69","IL2RA", "PTPRC", "SELL", "CCR7")
l_features <- sort(l_features)
```

```{r}
features <- c(m_features, l_features)
DotPlot(norm_so, features = features) + RotatedAxis()
```
The bubble plot is split between myloid and lymphoid features at AHR, AHR is the start of the lymphoid genes 

#### These features from from Tumor Infiltrating-immune cell human pathways image

```{r}
new_features <-c("CD83", "BST2", "HLA-DRA", "HLA-DRB5", "HLA-DRB1")
new_features <- sort(new_features)
```

Only new feature that this really added was BST2


### Creating Bubble Plot for Poster 
Here we reduce the number of features graphing to make our graphic for our poster simpler
```{r}
m_features_small <-c("HLA-DRA", "ITGAM", "ITGAX", "IL3RA","CD68", "CD14","CD86","FCGR3A","CD1C","GPR56","SIRPA","KIT","CD163","BST2","CD83")

m_features_small <- sort(m_features_small)

l_features_small <-c("CD3D","CD8A","CD4","CD19", "SDC1", "SPI1", "IFNG", "GATA3", "FOXP3", "TIFA", "CD69", "IL2RA", "PTPRC", "SELL", "CCR7")

l_features_small <- sort(l_features_small)

length(m_features_small) + length(l_features_small)

features_small <- c(m_features_small, l_features_small)
DotPlot(norm_so, features = features_small) + RotatedAxis()
```


### Renaming clusters for the umap labeling 
```{r}
# syntax to follow 
# norm_so$celltype[so$seurat_clusters == 1] <- 'CD8 Tcell'

norm_so$celltype[norm_so$seurat_clusters == 0] <- 'Cytotoxic T Cell'
norm_so$celltype[norm_so$seurat_clusters == 1] <- 'Cytotoxic T Cell'
norm_so$celltype[norm_so$seurat_clusters == 2] <- 'NK Cell'
norm_so$celltype[norm_so$seurat_clusters == 3] <- 'Macrophage'
norm_so$celltype[norm_so$seurat_clusters == 4] <- 'Cytotoxic T Cell'
norm_so$celltype[norm_so$seurat_clusters == 5] <- 'cDC'
norm_so$celltype[norm_so$seurat_clusters == 6] <- 'Regulatory T Cell'
norm_so$celltype[norm_so$seurat_clusters == 7] <- 'Cytotoxic T Cell'
norm_so$celltype[norm_so$seurat_clusters == 8] <- 'cDC'
norm_so$celltype[norm_so$seurat_clusters == 9] <- 'Macrophage'
norm_so$celltype[norm_so$seurat_clusters == 10] <- 'Plasma'
norm_so$celltype[norm_so$seurat_clusters == 11] <- 'Plasma'
norm_so$celltype[norm_so$seurat_clusters == 12] <- 'B Cell'
norm_so$celltype[norm_so$seurat_clusters == 13] <- 'Cytotoxic T Cell'
norm_so$celltype[norm_so$seurat_clusters == 14] <- 'Type I NK Cell'
norm_so$celltype[norm_so$seurat_clusters == 15] <- 'TBD'
norm_so$celltype[norm_so$seurat_clusters == 16] <- 'cDC2'
norm_so$celltype[norm_so$seurat_clusters == 17] <- 'Plasmacytoid Dendritic Cell'
norm_so$celltype[norm_so$seurat_clusters == 18] <- 'Macrophage'
norm_so$celltype[norm_so$seurat_clusters == 19] <- 'Polymorphonuclear MDSC'
norm_so$celltype[norm_so$seurat_clusters == 20] <- 'Plasmacytoid Dendritic Cell'
norm_so$celltype[norm_so$seurat_clusters == 21] <- 'Plasma'

```


### Running UMAP with new labeling 
```{r}
norm_so <- RunUMAP(norm_so, dims = 1:25)
DimPlot(norm_so, reduction = "umap", group.by = "celltype")
```
